Exploring how changes in age at death and cause of death have narrowed the black-white life expectancy gap over forty-five years in the United States.

### 2016-12-15 Manuscript/Manuscript.Rmd
added figures for females

```{r setup_figure1_and_2, echo=F}
year1 <- 1969
year2 <- 2013
sex <- "Male"

@@

for (state.i in levels(BlackWhite$State2)) {
  first.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year1 & paired.ids$Sex2 == sex)
  second.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year2 & paired.ids$Sex2 == sex)

@@

    temp2 <- data.frame(State2 = state.i,
                        contribution.to.gap.change(type.of.decomp = "COD",
                                                   list.cod.marginal.tables.smoothed[[first.index]],
                                                   list.cod.marginal.tables.smoothed[[second.index]])
    )

    temp3 <- data.frame(State2 = state.i,
                        contribution.to.gap.change(type.of.decomp = "Age",
                                                   list.age.decomp.tables.smoothed[[first.index]],
                                                   list.age.decomp.tables.smoothed[[second.index]])
    )

    cod.df <- rbind(cod.df, temp2)
    age.df <- rbind(age.df, temp3)
  }
}

##### COD ####
cod.df <- make_dataset_cod_plot(cod.decomp.table = cod.df, age.groups = "State2", 
                                 cause.of.death = "COD", sign.var = "narrowed_gap", 
                                 decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")

cod.df <- merge(cod.df, bw_dat, by = "State2")

@@

**Figure 1: Decomposition of the change in the black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
```{r figure1, fig.align = "center", fig.height=8, fig.width=8, echo=F}
x.labels <- levels(factor(cod.df$State.gdiff.order))
x.num.breaks <- length(levels(factor(cod.df$State.gdiff.order)))

@@

@@

**Figure 2: Decomposition of the change in the black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
```{r figure2, fig.align = "center", fig.height=8, fig.width=9, echo=F}
ggplotly(                                     
  ggplot(age.df, aes(y = State.gdiff.num, x = new.start)) +
    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = Ages), color = "white") + 
    geom_segment(data = bw_dat, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) + 
    geom_segment(data = bw_dat, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) + 
    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks, labels = x.labels) + scale_fill_viridis(discrete = T, direction = -1)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed")
  )
```

>>>>

```{r setup_figure1_and_3_males, echo=F}
year1 <- 1969
year2 <- 2013
sex <- "Male"

@@

for (state.i in levels(BlackWhite$State2)) {

  first.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year1 & paired.ids$Sex2 == sex)
  second.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year2 & paired.ids$Sex2 == sex)

@@

    temp2 <- data.frame(State2 = state.i, Sex2 = sex,
                        contribution.to.gap.change(type.of.decomp = "COD",
                                                   list.cod.marginal.tables.smoothed[[first.index]],
                                                   list.cod.marginal.tables.smoothed[[second.index]])
    )

    temp3 <- data.frame(State2 = state.i, Sex2 = sex,
                        contribution.to.gap.change(type.of.decomp = "Age",
                                                   list.age.decomp.tables.smoothed[[first.index]],
                                                   list.age.decomp.tables.smoothed[[second.index]])
    )

    cod.df <- rbind(cod.df, temp2)
    age.df <- rbind(age.df, temp3)
  }

}

##### COD ####
cod.df<- make_dataset_cod_plot(cod.decomp.table = cod.df, age.groups = "State2", 
                                 cause.of.death = "COD", sign.var = "narrowed_gap", 
                                 decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")

cod.df <- merge(cod.df, bw_dat, by = c("State2"))

@@

```{r setup_figure1_and_3_females, echo=F}
sex <- "Female"

bw_dat.f <- BlackWhite %>% 
  filter(Year3 %in% c(year1, year2), Sex2 == sex, is.na(WBgap.s) == F) %>%
  select(State2, Year3, WBgap.s) %>%
  arrange(Year3)

bw_dat.f <- tidyr::spread(bw_dat.f, Year3, WBgap.s) 
names(bw_dat.f)[2:3] <- c("first.gap","second.gap")  
bw_dat.f <- bw_dat.f %>% mutate(gap.diff = first.gap - second.gap,
                            State.gdiff.order = reorder(State2, first.gap),
                            State.gdiff.num = as.numeric(State.gdiff.order)
)

cod.df.f <- data.frame(State2 = factor(), COD = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())
age.df.f <- data.frame(State2 = factor(), Ages = factor(), Contribution.to.change = numeric(), Contrib.to.change.prop = numeric(), narrowed_gap = logical())

for (state.i in levels(BlackWhite$State2)) {

  first.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year1 & paired.ids$Sex2 == sex)
  second.index = which(paired.ids$State2 ==  state.i & paired.ids$Year3 == year2 & paired.ids$Sex2 == sex)

    temp2 <- data.frame(State2 = state.i, Sex2 = sex,
                        contribution.to.gap.change(type.of.decomp = "COD",
                                                   list.cod.marginal.tables.smoothed[[first.index]],
                                                   list.cod.marginal.tables.smoothed[[second.index]])
    )

    temp3 <- data.frame(State2 = state.i, Sex2 = sex,
                        contribution.to.gap.change(type.of.decomp = "Age",
                                                   list.age.decomp.tables.smoothed[[first.index]],
                                                   list.age.decomp.tables.smoothed[[second.index]])
    )

    cod.df.f <- rbind(cod.df.f, temp2)
    age.df.f <- rbind(age.df.f, temp3)
  }

}

##### COD ####
cod.df.f <- make_dataset_cod_plot(cod.decomp.table = cod.df.f, age.groups = "State2", 
                                 cause.of.death = "COD", sign.var = "narrowed_gap", 
                                 decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")

cod.df.f <- merge(cod.df.f, bw_dat.f, by = c("State2"))

cod.df.f <- cod.df.f %>% mutate(new.start = -start + first.gap, new.finish = -finish + first.gap)

#### Age ####
age.df.f <- make_dataset_cod_plot(cod.decomp.table = age.df.f, age.groups = "State2", 
                                  cause.of.death = "Ages", sign.var = "narrowed_gap", 
                                  decomp.var = "Contribution.to.change", decomp.var.prop = "Contrib.to.change.prop")

age.df.f <- merge(age.df.f, bw_dat.f, by = "State2")

age.df.f <- age.df.f %>% mutate(new.start = -start + first.gap, new.finish = -finish + first.gap)
```

Update the below text to also represent women:
**Figure 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs. their gap in 2013 (solid dashed line), with the change in the gap being represented by the number of years between these lines. The decomposition (in years) of the change in the gap by cause of death is also shown for each state. In the US, X of Y years of the change in gap is attributable to injuries, implying relatively more favourable *changes* to injury-related mortality rates in blacks vs. whites. The next large contributor to reducing the overall gap was "All other causes", which largely represents infant mortality and maternal mortality in our analysis (CR to confirm this). Blacks also saw relatively favourable changes in death related to communicable diseases, wher X of Y years of the change in the gap is attributable to changes in communicable disease-related mortality. Overall, X of Y years of the gap is attributable to cardiovarscular disease, but the story is very different by state. For example, California and Washington both underwent large unfavourable changes to CVD-related mortality among blacks relative to whites, where these conditions served to limit the change in the gap in California (which would have been much larger had CVD-related mortality rates had changed more favourably) and led to a much larger gap in the District of Columbia in 2013 vs. 1969, where X of Y years of the increase in the gap is attributable to CVD-related mortality.

**Figure 1: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
```{r figure1, fig.align = "center", fig.height=8, fig.width=8, echo=F}
x.labels <- levels(factor(cod.df$State.gdiff.order))
x.num.breaks <- length(levels(factor(cod.df$State.gdiff.order)))

@@

```
**This figure depicts how the gap in life expectancy for Males changed between 1969 (dashed vertical line) and 2013 (solid vertical line). Causes to the left of the dashed line narrowed the gap over time, whereas causes to the right exacerbated it.**

**Figure 2: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
```{r figure2, fig.align = "center", fig.height=8, fig.width=8, echo=F}
x.labels.f <- levels(factor(cod.df.f$State.gdiff.order))
x.num.breaks.f <- length(levels(factor(cod.df.f$State.gdiff.order)))

ggplotly(
  ggplot(cod.df.f, aes(y = State.gdiff.num, x = new.start)) +
    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, 
                  ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = COD), color = "white") +
    geom_segment(data = bw_dat.f, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) +
    geom_segment(data = bw_dat.f, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks.f, labels = x.labels.f)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed"))

@@

**Figure 3: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
```{r figure3, fig.align = "center", fig.height=8, fig.width=9, echo=F}
ggplotly(                                     
  ggplot(age.df, aes(y = State.gdiff.num, x = new.start)) +
    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = Ages), color = "white") + 
    geom_segment(data = bw_dat, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) + 
    geom_segment(data = bw_dat, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) + 
    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks, labels = x.labels) + scale_fill_viridis(discrete = T, direction = -1)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed")
  )
```

**Figure 4: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 2013 by US state**
```{r figure4, fig.align = "center", fig.height=8, fig.width=9, echo=F}
ggplotly(                                     
  ggplot(age.df.f, aes(y = State.gdiff.num, x = new.start)) +
    geom_rect(aes(xmin = new.start, ymin = State.gdiff.num - 0.45, ymax = State.gdiff.num + 0.45, xmax = new.finish, fill = Ages), color = "white") + 
    geom_segment(data = bw_dat.f, aes(x = second.gap, xend= second.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5)) + 
    geom_segment(data = bw_dat.f, aes(x = first.gap, xend= first.gap, y = State.gdiff.num - 0.5, yend = State.gdiff.num + 0.5), lty = 3) + 
    theme_minimal() + scale_y_continuous(breaks = 1:x.num.breaks.f, labels = x.labels.f) + scale_fill_viridis(discrete = T, direction = -1)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years)"), yaxis = list(title = NA, autorange = "reversed")
  )
```
